hatchR: A toolset to predict hatch and emergence phenology in wild fishes

### 2024-12-05 inst/manuscript/paper.Rmd
add species custom models and section on data checks and summary

@@

summarize_temp, check_continuous and plot_check_temp

@@

fit_model for three non-salmonid species

```{r, eval=FALSE}
smallmouth <- matrix(NA, 10, 2) |> tibble()
colnames(smallmouth) <- c("hours", "temp_F")
smallmouth$hours <- c(52, 54, 70, 78, 90, 98, 150, 167, 238, 234)
smallmouth$temp_F <- c(77, 75, 71, 70, 67, 65, 60, 59, 55, 55)

smallmouth <- smallmouth |> 
  mutate(days = ceiling(hours/24),
         temp_C = (temp_F -32) * (5/9))

plot(smallmouth$days~smallmouth$temp_C)

smb_mod <- fit_model(smallmouth,
                     temp = smallmouth$temp_C,
                     days = smallmouth$hours)

@@

>>>>

```{r echo=FALSE}
### 
library(tidyverse); library(hatchR)
```

@@

We demonstrate the utility of the check summary and check functions `summarize_temp()`, `plot_check_temp()`, and `check_continuous()` using a simulated year long dataset with temperature readings every thirty minutes.

```{r}
# create date object for a year with 30 min reading intervals
dates <- seq(from = ymd_hms("2000-07-01 00:00:00"),
             to = ymd_hms("2001-07-30 23:59:59"), length.out = 17568) 
# create empty dataframe
year_sim <- data.frame(matrix(NA, nrow = length(dates), ncol = 1))

# date column
colnames(year_sim)[1] <- "date"

# add dates vector to date column
year_sim[1] <- dates

#random seed
set.seed(123)

# take temps from a random normal dist with mean 10 sd 3 for every date time combo in dates and append to column (temp) in year_sim
year_sim$temp <- rnorm(n = length(dates), mean = 10, sd = 3) %>%
  abs()

dim(year_sim)

First, we recommend checking import data for any outliers or strange inputs using `plot_check_temp()`

```{r}
plot_check_temp(data = year_sim,
                dates = date,
                temperature = temp,
                temp_min = 0, # temp_min and max lines are user customizable
                temp_max = 25)
```

It appears there are no outliers but since each day has 48 records, we need to summarize it to daily mean temperature with `summarize_temp()` and then check for missing days with `check_continuous()`. We also recommend using `plot_check_temp()` again on the summarized data.

```{r}
# summarize
year_sim_summ <- summarize_temp(data = year_sim,
                                dates = date,
                                temperature = temp)

dim(year_sim_summ)

# check continuous
check_continuous(data = year_sim_summ,
                 dates = date)

# we can demonstrate an error by removing the Oct 8 (100th day)
check_continuous(data = year_sim_summ[-100,],
                 dates = date)

# it is useful to plot again to check your summarized data
plot_check_temp(data = year_sim_summ,
                dates = date,
                temperature = daily_temp,
                temp_min = 0, # temp_min and max lines are user customizable
                temp_max = 15)
```

@@

fit_model for three non-salmonid specieslib

Below, we demonstrate how the `fit_model()` function may be used to create custom parameterizations for species beyond the Salmonids in the `model_table` included in the package. Furthermore, we include parameterizations from three warm-water species to demonstrate the `fit_model()` utility for species beyond the original scope of the original effective value approach. We demonstrate parameterizations for commonly cultured sportfishes including Smallmouth Bass (*Micropterus dolomieu*), Channel Catfish (*Ictalurus punctatus*) from @small2001, and Lake Sturgeon (*Acipenser fulvescens*) from @smith2005.

We demonstrate the utility of this approach by creating a random thermal regime with an ascending thermograph with a mean temperature of 16 °C, parameterizing models for each species, and demonstrating days to hatch and developmental period for each species with the random thermal regime.

```{r, eval=FALSE}
###  make temp regime
set.seed(1234)
# create random temps and corresponding dates
temps_sim <- sort(rnorm(n =30, mean = 16, sd = 1), decreasing = FALSE)
dates_sim <-  seq(from = ymd("2000-07-01"),
             to = ymd("2000-07-31"), length.out = 30)

data_sim <- cbind (dates_sim, temps_sim) |> data.frame()

### smallmouth mod
smallmouth <- matrix(NA, 10, 2) |> tibble()
colnames(smallmouth) <- c("hours", "temp_F")
smallmouth$hours <- c(52, 54, 70, 78, 90, 98, 150, 167, 238, 234)
smallmouth$temp_F <- c(77, 75, 71, 70, 67, 65, 60, 59, 55, 55)

# change F to C and hours to days
smallmouth <- smallmouth |> 
  mutate(days = ceiling(hours/24),
         temp_C = (temp_F -32) * (5/9))

smb_mod <- fit_model(temp = smallmouth$temp_C,
                     days = smallmouth$days,
                     species = "smb",
                     dev.type = "hatch")

### catfish mod
catfish <- matrix(NA, 3, 2) |> tibble()
colnames(catfish) <- c("days", "temp_C")
catfish$days <- c(16,21,26)
catfish$temp_C <- c(22,10,7)

cat_mod <- fit_model(temp = catfish$temp_C,
                     days = catfish$days,
                     species = "catfish",
                     dev.type = "hatch")

### lake sturgeon mod
sturgeon <-  matrix(NA, 7, 2) |> tibble()
colnames(sturgeon) <- c("days", "CTU")
sturgeon$days <- c(7,5,6,6,5,11,7)
sturgeon$CTU <- c(58.1, 62.2, 61.1, 57.5, 58.1, 71.4, 54.7)

sturgeon <- sturgeon |> 
  mutate(temp_C = CTU/days)

plot(sturgeon$days ~ sturgeon$temp_C)

sturgeon_mod <- fit_model(days = sturgeon$days,
                          temp = sturgeon$temp_C,
                          species = "sturgeon",
                          dev.type = "hatch")

### predict_phenology

#smallmouth bass
smb_res <- predict_phenology(data = data_sim,
                  dates = dates_sim,
                  temperature = temps_sim,
                  spawn.date = "2000-07-01",
                  model = smb_mod$func)

# catfish
catfish_res <- predict_phenology(data = data_sim,
                  dates = dates_sim,
                  temperature = temps_sim,
                  spawn.date = "2000-07-01",
                  model = cat_mod$func)

# sturgeon
# note that 16 C is pretty far out of range of temps for model fit, not best practice
sturgeon_res <- predict_phenology(data = data_sim,
                  dates = dates_sim,
                  temperature = temps_sim,
                  spawn.date = "2000-07-01",
                  model = sturgeon_mod$func)

@@

```{r, echo=FALSE}
# summary for all species
all_res <- data.frame(matrix(NA, 3, 2))
colnames(all_res) <- c("start", "stop")
all_res[1,] <- smb_res$dev.period
all_res[2,] <- catfish_res$dev.period
all_res[3,] <- sturgeon_res$dev.period

all_res <- all_res |> 
  mutate(days = stop-start)
all_res$Species <- c("Smallmouth Bass", "Channel Catfish", "Lake Sturgeon")
```
